From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-20 paper/paper.Rmd
wip

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: after all, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticty literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; species traits
and the interaction of ecological and evolutionary processes [@Ozgul2009; @Lenormand2009; @Vindesens2015; @Schreiber2015],
not to mention the extensive literature in stochastic evolution and popluation genetics
[e.g. Ewens, @Lewontin1969]

and the role of stochasticity in more complex models, particularly in structured populations (matrix models),

in

has been spent on illustrating the importance that complexity: age and stage
structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
While acknowledging more realistic examples and more general results where avialable,
using specific and familiar simple models that -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

Few topics have recieved greater attention then the question of
the persistence of a given species [@Kendall1949;  @Bartlett1959; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]

>>>>

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: after all, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticty literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009; @Lenormand2009; @Vindenes2015; @Schreiber2015],
not to mention the extensive literature in stochastic evolution and popluation genetics
[e.g. @Ewens2004, @Lewontin1969].

While acknowledging more realistic examples and more general results where available,
using specific and familiar simple models -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

Few topics have recieved greater attention then the question of
the persistence of a given species [@Kendall1949;  @Bartlett1955; @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
